An in-depth study of Explainability and Adversarial Robustness for RNNs

### 2019-12-07 ex3/paper/paper.tex
Add some more feature correlation

Most features such as, e.g., TCP flags cannot be easily manipulated because it might violate the TCP and thus cause communication to fail. Hence, we experiment whether by manipulating the interarrival times and packet lengths of the first packets, the classifier might be fooled. Our scenario is quite different from how generating adversarial samples usually works: We have significantly less features than in, for example, an image. Furthermore, only two out of those few features can be manipulated. And even these features cannot be manipulated freely: Besides the constraints that packet size and interarrival time cannot become negative there are also more problem-specific constraints:
\begin{itemize}
\item Only packets can be manipulated which are transmitted by the attacker. This means the attacker can only control one direction (except for botnets where both sides are attackers).
\item Packets must not be smaller than initially, as otherwise less information could be transmitted or packets might be rendered syntactically invalid.
\item Interarrival times must not decrease, as otherwise the physical speed of data transmission can be violated in some cases. Since an in-depth analysis of when the reduction of interarrival times is possible, would come with substantial effort, we generally disallowed a reduction of this feature.
\end{itemize}

@@

\subsection{Evaluation}
% \begin{figure*}
% \includegraphics[width=0.45\textwidth]{adv_plots/cwl1.pdf}
% \includegraphics[width=0.45\textwidth]{adv_plots/cwl2.pdf}
% \caption{Performance of the \gls{cw} with different $\epsilon$ values for $\ell_1$ and $\ell_2$ distances.}
% \label{fig:adv_cw}
% \end{figure*}
\autoref{fig:adv_cw} depicts the ratio of attack flows which were successfully misclassified as benign with respect to the final packet along with the resulting distance from the original sample. Evidently, the larger the $\epsilon$ parameter, the better the attack works, but also the distances become larger.

@@

\begin{table*}
\caption{Attack success ratios per attack family. Depicted are flow detection accuracies for adversarial flows (left numbers) and flow detection accuracies for unmodified flows (right numbers).}
\label{tab:adv_per_family}
\centering
\begin{tabular}{lllllllll}
\toprule
Attack type & FGSM & Carlini-Wagner, $\ell_1$ & Carlini-Wagner, $\ell_2$ & $\ell_\infty$-bounded \gls{pgd} \\
\midrule
Botnet:ARES	&	0.0000 / 0.0000	&	0.0000 / 0.0000	&	0.0000 / 0.0000	&	0.0000 / 0.0000	\\
Brute Force:FTP-Patator	&	0.2800 / 0.5800	&	0.1600 / 0.5800	&	0.1600 / 0.5800	&	0.2800 / 0.5800	\\
Brute Force:SSH-Patator	&	0.8300 / 1.0000	&	0.0000 / 1.0000	&	0.0000 / 1.0000	&	0.8900 / 1.0000	\\
DDoS:LOIT	&	0.8300 / 0.8600	&	0.6400 / 0.8600	&	0.5600 / 0.8600	&	0.8200 / 0.8600	\\
DoS / DDoS:DoS GoldenEye	&	0.8900 / 1.0000	&	0.5000 / 1.0000	&	0.4200 / 1.0000	&	0.9200 / 1.0000	\\
DoS / DDoS:DoS Hulk	&	0.6700 / 1.0000	&	0.5400 / 1.0000	&	0.4700 / 1.0000	&	1.0000 / 1.0000	\\
DoS / DDoS:DoS Slowhttptest	&	0.8700 / 1.0000	&	0.4100 / 1.0000	&	0.1900 / 1.0000	&	0.7400 / 1.0000	\\
DoS / DDoS:DoS slowloris	&	0.4400 / 0.9700	&	0.3100 / 0.9700	&	0.1300 / 0.9700	&	0.9200 / 0.9700	\\
DoS / DDoS:Heartbleed	&	0.0000 / 0.0000	&	0.0000 / 0.0000	&	0.0000 / 0.0000	&	0.0000 / 0.0000	\\
Infiltration:Dropbox download 	&	0.6100 / 0.9600	&	0.4300 / 0.9600	&	0.3800 / 0.9600	&	0.7400 / 0.9600	\\
PortScan:PortScan - Firewall off	&	0.6800 / 0.9100	&	0.3500 / 0.9100	&	0.3500 / 0.9100	&	0.7200 / 0.9100	\\
PortScan:PortScan - Firewall on	&	0.7400 / 0.8700	&	0.6100 / 0.8700	&	0.5300 / 0.8700	&	0.7300 / 0.8700	\\
Web Attack:Sql Injection	&	0.0000 / 0.6923	&	0.0000 / 0.6923	&	0.0000 / 0.6923	&	0.0000 / 0.6923	\\
Web Attack:XSS	&	0.6700 / 0.9500	&	0.7200 / 0.9500	&	0.6200 / 0.9500	&	0.6700 / 0.9500	\\
\bottomrule

@@

\begin{itemize}
\item
\textbf{Feature quantity.}
The number of features that is fed into an \gls{rnn} is found as the product of packet features with the total length of the flow. For long flows, the total number of inputs can therefore grow to a substantial size.

@@

\item
\textbf{Multiple prediction outputs.}
Often an \gls{rnn} produces an output at each time step. When trying to apply explainability methods, hence, the question arises which output to base the methods on.

The natural choice is to base the methods on the prediction output which occurs at the same time step as the feature under investigation. Not only does this approach require a substantially lower amount of computational resources compared to predictions occurring at a later time step, but we also expect the immediately occurring prediction outcome to exhibit a stronger correlation to the feature under investigation, compared to a prediction outcome occurring many time steps later.
However, due to the complex decision processes that can occur for deep neural networks, it is absolutely  uncertain if the network's final prediction is correctly depicted when following this approach.
\end{itemize}

@@

% numerator = max(float(baseline_accuracy-accuracy_for_feature_combination), 0)
% 			denominator = max(float(baseline_accuracy-accuracy_for_feature_1), 0) + max(float(baseline_accuracy-accuracy_for_feature_2), 0)
\begin{equation}
\text{score} = \frac{\text{acc}_\text{base}-\text{acc}_\text{-both\_features}}{\left(\text{acc}_\text{base}-\text{acc}_\text{-feature\_1}\right) + \left(\text{acc}_\text{base}-\text{acc}_\text{-feature\_2}\right)}
\end{equation}
With $\text{acc}_\text{base}$ we denote the average accuracy of the classifier with all features included, with $\text{acc}_\text{-feature\_i}$ the average accuracy if feature $i$ is omitted and with $\text{acc}_\text{-both\_features}$ the average accuracy if both features are omitted. The higher the resulting number, the larger the information that both features share. If the score is one, this means that both features contain no joint information.

\subsubsection{Mutual Information}
\begin{figure}
\includegraphics[width=\columnwidth]{../plots/mutinfo_example.pdf}
\caption{Distributions $f_A(x)$ yielding an identical accuracy drop.}
\label{fig:mutinfo_example}
\end{figure}
Input Perturbation has some shortcomings when considering feature sensitivity. Assuming that the test dataset is representative for production use, for feature importance it is reasonable to equate the distribution of the perturbed feature with the feature distribution. On the other hand, when evaluating feature sensitivity, e.g. for analyzing potential for adversarial samples, the attacker is not limited by this distribution and frequently is able to choose arbitrary values in the feature space.

Furthermore, let $f_A(x)$  denote the joint probability for classification as attack with a feature value $x$. Accuracy then is $\int f_A(x) dx$ for attacks and $1-\int f_A(x) dx$ for benign traffic. Consider \autoref{fig:mutinfo_example}. Evidently, both distributions
yield the same accuracy, but clearly the right-side distribution has a larger influence on the prediction, as in the right-side case the prediction deterministically depends on the feature value.
To capture such dependencies, we propose to use mutual information instead of accuracy drop to determine feature sensitivity.
Mutual Information between two random variables $X,Y$ is defined as 
\begin{equation}
I_{X,Y} = \mathbb E \left\{ \log\left(\frac{f_{X,Y}(x,y)}{f_X(x)f_Y(y)}\right) \right\},
\end{equation}
with $f_X(x), f_Y(y)$ and $f_{X,Y}(x,y)$ denoting the distribution of $X,Y$ and their joint distribution, respectively. To obtain feature sensitivity, we compute mutual information between an input variable and the prediction output for one flow at one particular time step, and average over the result for the test dataset.

@@

\subsection{Reducing features}
The most obvious defence, which is only possible because of the nature of the dataset, is to simply leave out features which the attacker can manipulate. For this, we try two different approaches:
\begin{itemize}
\item Leaving out all features that are manipulable: packet size and interarrival time
\item Leaving out manipulable features when the attacker can manipulate them: only in the direction from the attacker to the victim (This, however does not prevent adversarial samples for botnets, for which both sides are malicious)
\end{itemize}

>>>>

Most features such as, e.g., TCP flags cannot be easily manipulated because it might violate the TCP and thus cause communication to fail. Hence, we experiment whether by manipulating the interarrival times and packet lengths of the first packets, the classifier might be fooled. Our scenario is quite different from how generating adversarial samples usually works: We have significantly less features than in, for example, an image. Furthermore, only two out of those few features can be manipulated. And even these features cannot be manipulated freely: Besides the constraints that packet size and interarrival time cannot become negative there are also more problem-specific constraints:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Only packets can be manipulated which are transmitted by the attacker. This means the attacker can only control one direction (except for botnets where both sides are attackers).
\item Packets must not be smaller than initially, as otherwise less information could be transmitted or packets might be rendered syntactically invalid.
\item Interarrival times must not decrease, as otherwise the physical speed of data transmission can be violated in some cases. Since an in-depth analysis of when the reduction of interarrival times is possible, would come with substantial effort, we generally disallowed a reduction of this feature.
\end{itemize}

@@

\subsection{Evaluation}
% \begin{figure*}
% \includegraphics[width=0.45\textwidth]{adv_plots/cwl1.pdf}
% \includegraphics[width=0.45\textwidth]{adv_plots/cwl2.pdf}
% \caption{Performance of the \gls{cw} with different $\epsilon$ values for $\ell_1$ and $\ell_2$ distances.}
% \label{fig:adv_cw}
% \end{figure*}
\todo{Max: Fix me. Don't know how the reference became invalid...}
\autoref{fig:adv_cw} depicts the ratio of attack flows which were successfully misclassified as benign with respect to the final packet along with the resulting distance from the original sample. Evidently, the larger the $\epsilon$ parameter, the better the attack works, but also the distances become larger.

@@

\begin{table*}
\caption{Attack success ratios per attack family. Depicted are flow detection accuracies for adversarial flows (left numbers) and flow detection accuracies for unmodified flows (right numbers).}
\label{tab:adv_per_family}
\centering
\begin{tabular}{llllllllll}
\toprule
Attack type & \multicolumn{2}{l}{FGSM} & \multicolumn{2}{l}{Carlini-Wagner, $\ell_1$} & \multicolumn{2}{l}{Carlini-Wagner, $\ell_2$} & \multicolumn{2}{l}{$\ell_\infty$-bounded \gls{pgd}} \\
\cmidrule(r){2-3} \cmidrule(lr){4-5} \cmidrule(l){6-7} \cmidrule(l){8-9}
& adv. & orig. & adv. & orig. & adv. & orig. & adv. & orig. \\
\midrule
Botnet:ARES	&	0.0000 & 0.0000	&	0.0000 & 0.0000	&	0.0000 & 0.0000	&	0.0000 & 0.0000	\\
Brute Force:FTP-Patator	&	0.2800 & 0.5800	&	0.1600 & 0.5800	&	0.1600 & 0.5800	&	0.2800 & 0.5800	\\
Brute Force:SSH-Patator	&	0.8300 & 1.0000	&	0.0000 & 1.0000	&	0.0000 & 1.0000	&	0.8900 & 1.0000	\\
DDoS:LOIT	&	0.8300 & 0.8600	&	0.6400 & 0.8600	&	0.5600 & 0.8600	&	0.8200 & 0.8600	\\
DoS / DDoS:DoS GoldenEye	&	0.8900 & 1.0000	&	0.5000 & 1.0000	&	0.4200 & 1.0000	&	0.9200 & 1.0000	\\
DoS / DDoS:DoS Hulk	&	0.6700 & 1.0000	&	0.5400 & 1.0000	&	0.4700 & 1.0000	&	1.0000 & 1.0000	\\
DoS / DDoS:DoS Slowhttptest	&	0.8700 & 1.0000	&	0.4100 & 1.0000	&	0.1900 & 1.0000	&	0.7400 & 1.0000	\\
DoS / DDoS:DoS slowloris	&	0.4400 & 0.9700	&	0.3100 & 0.9700	&	0.1300 & 0.9700	&	0.9200 & 0.9700	\\
DoS / DDoS:Heartbleed	&	0.0000 & 0.0000	&	0.0000 & 0.0000	&	0.0000 & 0.0000	&	0.0000 & 0.0000	\\
Infiltration:Dropbox download 	&	0.6100 & 0.9600	&	0.4300 & 0.9600	&	0.3800 & 0.9600	&	0.7400 & 0.9600	\\
PortScan:PortScan - Firewall off	&	0.6800 & 0.9100	&	0.3500 & 0.9100	&	0.3500 & 0.9100	&	0.7200 & 0.9100	\\
PortScan:PortScan - Firewall on	&	0.7400 & 0.8700	&	0.6100 & 0.8700	&	0.5300 & 0.8700	&	0.7300 & 0.8700	\\
Web Attack:Sql Injection	&	0.0000 & 0.6923	&	0.0000 & 0.6923	&	0.0000 & 0.6923	&	0.0000 & 0.6923	\\
Web Attack:XSS	&	0.6700 & 0.9500	&	0.7200 & 0.9500	&	0.6200 & 0.9500	&	0.6700 & 0.9500	\\
\bottomrule

@@

\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item
\textbf{Feature quantity.}
The number of features that is fed into an \gls{rnn} is found as the product of packet features with the total length of the flow. For long flows, the total number of inputs can therefore grow to a substantial size.

@@

\item
\textbf{Multiple prediction outputs.}
Often an \gls{rnn} produces an output at each time step. When trying to apply explainability methods, hence, the question arises which output to base the methods on. The natural choice is to base the methods on the prediction output which occurs at the same time step as the feature under investigation. Not only does this approach require a substantially lower amount of computational resources compared to predictions occurring at a later time step, but we also expect the immediately occurring prediction outcome to exhibit a stronger correlation to the feature under investigation, compared to a prediction outcome occurring many time steps later.
However, due to the complex decision processes that can occur for deep neural networks, it is absolutely  uncertain if the network's final prediction is correctly depicted when following this approach.
\end{itemize}

@@

\begin{equation}
\text{score} = \frac{\text{acc}_\text{base}-\text{acc}_\text{-both\_features}}{\left(\text{acc}_\text{base}-\text{acc}_\text{-feature\_1}\right) + \left(\text{acc}_\text{base}-\text{acc}_\text{-feature\_2}\right)}
\end{equation}
With $\text{acc}_\text{base}$ we denote the average accuracy of the classifier with all features included, with $\text{acc}_\text{-feature\_i}$ the average accuracy if feature $i$ is omitted and with $\text{acc}_\text{-both\_features}$ the average accuracy if both features are omitted. The higher the resulting number, the larger the information that both features share. If the score is one, this means that both features contain no joint information. For example, the score between RST and the protocol identifier is 8.5; the highest of all pairs of features. While this may seem unintuitive at the first glance, it likely stems from the fact that if the protocol identifier (TCP or UDP) is missing, then if RST is 1 at some point of the flow, the flow has to be TCP.

\subsubsection{Mutual Information}
\begin{figure}
\includegraphics[width=\columnwidth]{../plots/mutinfo_example.pdf}
\caption{Two distributions yielding an identical accuracy drop.}
\label{fig:mutinfo_example}
\end{figure}
Input Perturbation has some shortcomings when considering feature sensitivity. Assuming that the test dataset is representative for production use, for feature importance it is reasonable to equate the distribution of the perturbed feature with the feature distribution. On the other hand, when evaluating feature sensitivity, e.g. for analyzing potential for adversarial samples, the attacker is not limited by this distribution and frequently is able to choose arbitrary values in the feature space.

Furthermore, let $f_A(x)$  denote the joint probability \todo{Max: Ist joint probability nicht immer wenn man mehrere Ereignisse hat? Ich hätte hier probability density function geschrieben...} for classification as attack with a feature value $x$. Accuracy then is $\int f_A(x) dx$ for attacks and $1-\int f_A(x) dx$ for benign traffic. \autoref{fig:mutinfo_example} shows two different distributions
yielding the same accuracy, but clearly the right-side distribution has a larger influence on the prediction, as in the right-side case the prediction deterministically depends on the feature value.
To capture such dependencies, we propose to use mutual information instead of accuracy drop to determine feature sensitivity.
Mutual Information between two random variables $X,Y$ is defined as 
\begin{equation}
I_{X,Y} = \mathbb E \left\{ \log\left(\frac{f_{X,Y}(x,y)}{f_X(x)f_Y(y)}\right) \right\},
\end{equation}
with $f_X(x), f_Y(y)$ and $f_{X,Y}(x,y)$ denoting the distribution of $X,Y$ and their joint distribution, respectively. To obtain feature sensitivity, we compute mutual information between an input variable and the prediction output for one flow at one particular time step, and average over the result for the test dataset.

@@

\subsection{Reducing features}
The most obvious defence, which is only possible because of the nature of the dataset, is to simply leave out features which the attacker can manipulate. For this, we try two different approaches:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Leaving out all features that are manipulable: packet size and interarrival time
\item Leaving out manipulable features when the attacker can manipulate them: only in the direction from the attacker to the victim (This, however does not prevent adversarial samples for botnets, for which both sides are malicious)
\end{itemize}
